Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-08 rnni_convexity.tex
to fig nni_vs_rnni

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $\rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
We will give a counterexample to the Split Theorem in $\rnni$ and consider the Weak Split Theorem that states that every shared split of taxa if maintained along a shortest path (see Conjecture~\ref{conjecture:cluster_theorem}).

@@

\begin{proof}
Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
Let us assume that $T$ and $R = [C_1, \ldots, C_{n-2}]$ are trees for which $\findpath$ computes a path of maximum length.
It follows that in each step of $\findpath$ the most recent common ancestor of the cluster $C_k$ considered in that step is the root.
Thus, there are $n-1-k$ $\rnni$ operations needed to move $C_k$ to its correct position.
Hence, the maximum length of a path computed by $\findpath$ is bounded by $\sum\limits_{i = 1}^{n-2} n-1-i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

\begin{proof}
First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
So we can follow that an $\rnni$ move on $T$ can decrease the rank of $\parent_T(n+1)$ by at most one.

Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$ as the moves on $p$ that are left out to receive $p{\big|}_n$ are only ones involving the parent of taxon $n+1$, which do not affect the remaining $\rnni$ moves.

Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
With the knowledge that an $\rnni$ move on $T$ can change the rank of $\parent_T(n+1)$ by at most one, we can conclude that $|p{\big|}_n| \leq |p| - \delta$.
Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
\end{proof}

@@

In this section we restrict our attention to the set of caterpillar trees.
These trees are of particular interest because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space, which we will show in the following.
We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time, which underlines the importance of these trees once more and suggests that there is a difference in the complexity of computing distances in $\nni$ and $\rnni$.
Throughout this section we represent these trees with lists as described in Section~\ref{section:alg_csort}.

Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$ building a cherry and moving it may result a path that is shorter than a path where taxa move separately as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
In this example only one move is needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
In $\rnni$ space however, there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
As a consequence a path where taxa move separately and no new cherry is built is shortest in $\rnni$.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_vs_RNNI}
\vspace{12pt}
\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
The upper path is a natural extension of the shortest $\nni$ path in $\rnni$.
It is longer than a shortest $\rnni$ path as the one at the bottom that consists of caterpillar trees only.}
\label{fig:NNI_vs_RNNI}
\end{figure}

>>>>

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $\rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem that states that every shared split of taxa is maintained along a shortest path.
We will call this statement the Clade Theorem, as it is slightly more natural to work with ranked trees in terms clades rather than splits (see Conjecture~\ref{conjecture:cluster_theorem}).

@@

\begin{proof}
Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
Let us assume that $T$ and $R = [C_1, \ldots, C_{n-2}]$ are trees for which $\findpath$ computes a path of maximum length.
It follows that in each step of $\findpath$ the most recent common ancestor of the cluster $C_k$ considered in that step is the root.
Thus, there are $n-1-k$ $\rnni$ operations necessary to move $C_k$ to its correct position.
Hence, the maximum length of a path computed by $\findpath$ is bounded by $\sum\limits_{i = 1}^{n-2} (n-1-i) = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

\begin{proof}
First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
Hence an $\rnni$ move on $T$ can decrease the rank of $\parent_T(n+1)$ by at most one.

Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$.
Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
Combining this with the observations above, we conclude that $|p{\big|}_n| \leq |p| - \delta$.
Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
\end{proof}

@@

In this section we restrict our attention to the set of caterpillar trees.
These trees are of particular interest because they are used to prove the computing the $\nni$ distance is an NP-hard problem.
As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time.
Throughout this section we use the list representation of trees described in Section~\ref{section:alg_csort}.

Recall \autocite{Gavryushkin2018-ol} that to compute a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then move the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
In $\rnni$ however, the necessity of additional rank moves invalidates this strategy.
This is due to the fact that $\nni$ moves $\rnni$ are only allowed on edges of length one.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_vs_RNNI}
\vspace{12pt}
\caption{Paths between caterpillar trees $T$ and $R$: Solid arrows indicate paths in $\rnni$, dashed arrow is a move only possible in $\nni$.
Bottom path is a shortest $\rnni$ path.}
\label{fig:NNI_vs_RNNI}
\end{figure}
